---
title: "Practical 8: Natural Language Processing"
author: "Ward B. Eiling"
format: 
  html:
    self-contained: true
execute:
  warning: false
  message: false
---

# Introduction

In this practical, we will learn word embeddings to represent text data, and we will also analyse a recurrent neural network.

We use the following packages:

```{r}
library(magrittr)  # for pipes
library(tidyverse) # for tidy data and pipes
library(ggplot2)   # for visualization
library(wordcloud) # to create pretty word clouds
library(stringr)   # for regular expressions
library(text2vec)  # for word embedding
library(tidytext)  # for text mining
library(tensorflow)
library(keras3)
```

# Take-home exercises

## Word embedding

In the first part of the practical, we will apply word embedding approaches. A key idea in working with text data concerns representing words as numeric quantities. There are a number of ways to go about this as we reviewed in the lecture. Word embedding techniques such as word2vec and GloVe use neural networks approaches to construct word vectors. With these vector representations of words we can see how similar they are to each other, and also perform other tasks such as sentiment classification.

Let’s start the word embedding part with installing the `harrypotter` package using [devtools](https://www.r-project.org/nosvn/pandoc/devtools.html). The `harrypotter` package supplies the first seven novels in the Harry Potter series. You can install and load this package with the following code:

```{r}
# devtools::install_github("bradleyboehmke/harrypotter")
library(harrypotter) # Not to be confused with the CRAN palettes package
```

**1. Use the code below to load the first seven novels in the Harry Potter series.**

```{r}
hp_books <- c("philosophers_stone", "chamber_of_secrets",
              "prisoner_of_azkaban", "goblet_of_fire",
              "order_of_the_phoenix", "half_blood_prince",
              "deathly_hallows")

hp_words <- list(
  philosophers_stone,
  chamber_of_secrets,
  prisoner_of_azkaban,
  goblet_of_fire,
  order_of_the_phoenix,
  half_blood_prince,
  deathly_hallows
) %>%
  # name each list element
  set_names(hp_books) %>%
  # convert each book to a data frame and merge into a single data frame
  map_df(as_tibble, .id = "book") %>%
  # convert book to a factor
  mutate(book = factor(book, levels = hp_books)) %>%
  # remove empty chapters
  filter(!is.na(value)) %>%
  # create a chapter id column
  group_by(book) %>%
  mutate(chapter = row_number(book))

head(hp_words)
```

**2. Convert the hp_words object into a dataframe and use the `unnest_tokens()` function from the tidytext package to tokenize the dataframe.**

```{r}
hp_tokens <- as.data.frame(hp_words) %>%
  unnest_tokens(input = "value", output = "word", token = "words", to_lower = TRUE)
```

**3. Remove the stop words from the tokenized data frame.**

```{r}
hp_tokens2 <- hp_tokens %>%
  anti_join(stop_words, by = "word")
```

**4. Creates a vocabulary of unique terms using the create_vocabulary() function from the text2vec package and remove the words that they appear less than 5 times.**

```{r}
# create variable with book with chapter
ids <- hp_tokens2 %>%
  unite(book_chapter, book, chapter, sep = "_") %>%
  select(book_chapter) %>%
  pull()

it = itoken(list(hp_tokens2$word))
vocab = create_vocabulary(it)
vocab2 = prune_vocabulary(vocab, term_count_min = 5)
```

**5. The next step is to create a token co-occurrence matrix (TCM). The definition of whether two words occur together is arbitrary. First create a vocab_vectorizer, then use a window of 5 for context words to create the TCM.**

```{r}
# create a vocab vectorizer
vectorizer = vocab_vectorizer(vocab2)

# create a token co-occurrence matrix
it2 = itoken(vocab2$term)
hp_tcm = create_tcm(it, vectorizer, skip_grams_window = 5L)
```

**6. Use the GlobalVectors as given in the code below to fit the word vectors on our data set. Choose the embedding size (rank variable) equal to 50, and the maximum number of co-occurrences equal to 10. Train word vectors in 20 iterations. You can check the full input arguments of the fit_transform function from [here](https://www.rdocumentation.org/packages/text2vec/versions/0.5.1/topics/GlobalVectors).**

```{r}
glove <- GlobalVectors$new(rank = 50, x_max = 10)
hp_wv_main <- glove$fit_transform(hp_tcm, n_iter = 20, convergence_tol = 0.001)
dim(hp_wv_main)
```

**7. The GloVe model learns two sets of word vectors: main and context. Essentially they are the same since the model is symmetric. From the experience learning two sets of word vectors leads to higher quality embeddings (read more [here](http://text2vec.org/glove.html)). Best practice is to combine both the main word vectors and the context word vectors into one matrix. Extract the word vectors and save the summation of them for further questions.**

```{r}
hp_wv_context = glove$components
dim(hp_wv_context)
hp_wv_main_context = hp_wv_main + t(hp_wv_context)
dim(hp_wv_main_context)
```

**8. Find the most similar words to words “harry”, “death”, and “love”. Use the `sim2` function with the cosine similary measure.**

```{r}
# find the most similar words to "harry", "death", and "love"
sim_words <- sim2(hp_wv_main_context, method = "cosine")
sim_words_selected <- t(sim_words[rownames(sim_words) %in% c("harry", "death", "love"),])
# for each column selected the second lowest value (the lowest is the word itself)
sim_words_selected2 <- apply(sim_words_selected, 2, function(x) names(sort(x, decreasing = TRUE)[1:10]))
sim_words_selected2
```

The most similar word to love is "potion", for death it is "eaters" and for harry it is "ron".

**9. Now you can play with word vectors! For example, add the word vector of “harry” with the word vector of “love” and subtract them from the word vector of “death”. What are the top terms in your result?**

```{r}
# Perform vector arithmetic
harry_love_minus_death <- hp_wv_main_context["harry",] + hp_wv_main_context["love",] - hp_wv_main_context["death",]

# Calculate cosine similarity between the result and all word vectors
similarity_result <- sim2(x = hp_wv_main_context, y = matrix(harry_love_minus_death, nrow = 1), method = "cosine")

# Sort the similarity scores in descending order
top_similar_words <- sort(similarity_result[, 1], decreasing = TRUE)

# Extract the top 10 most similar terms
top_terms <- names(top_similar_words)[1:10]
top_terms
```

The top terms are "hermione", "ron" and "harry". This is not surprising as these characters are close to Harry (so +love) and far from dead (-dead).

# Lab exercises

## Sentiment classification with RNN

For sentiment classification with pre-trained word vectors, we want to use [GloVe](https://nlp.stanford.edu/projects/glove/) pretrained word vectors. These word vectors were trained on Wikipedia 2014 and Gigaword 5 containing 6B tokens, 400K vocab, uncased, 50d, 100d, 200d, & 300d vectors. Download the `glove.6B.300d.txt` file manually from the website or use the code below for this purpose.

```{r}
# Download Glove vectors if necessary
# if (!file.exists('glove.6B.zip')) {
#   download.file('https://nlp.stanford.edu/data/glove.6B.zip',destfile = 'glove.6B.zip')
#   unzip('glove.6B.zip')
# }
```

**10. Use the code below to load the pre-trained word vectors from the file ‘glove.6B.300d.txt’ (if you have memory issues load the file ‘glove.6B.50d.txt’ instead).**

```{r}
# unzip the file
# unzip('practical_08_natural-language-processing/glove.6B.zip')

# load glove vectors
vectors <- data.table::fread('practical_08_natural-language-processing/glove.6B.300d.txt', data.table = F, encoding = 'UTF-8', quote = "")
colnames(vectors) <- c('word', paste('dim',1:300,sep = '_'))

# convert vectors to dataframe
vectors <- as_tibble(vectors)
```

**11. IMDB movie reviews is a labeled data set available with the `text2vec` package. This data set consists of 5000 IMDB movie reviews, specially selected for sentiment analysis. The sentiment of the reviews is binary, meaning an IMDB rating \< 5 results in a sentiment score of 0, and a rating \>=7 has a sentiment score of 1. No individual movie has more than 30 reviews. Load this data set and convert it to a dataframe.**

```{r}
movie_review <- as.data.frame(text2vec::movie_review)
```

**12. To create a learning model using `Keras`, let’s first define the hyperparameters. Define the parameters of your `Keras` model with a maximum of 10000 words, maxlen of 60 and word embedding size of 300 (if you had memory problems change the embedding dimension to a smaller value, e.g., 50).**

```{r}
# define hyperparameters
max_words <- 10000
maxlen <- 60
dim_size <- 300
```

**13. Use the text_tokenizer function from `Keras` and tokenize the imdb review data using a maximum of 10000 words.**

```{r}
library(keras) # keras3 doesn't have text_tokenizer

# tokenize the imdb review data
tokenizer <- keras::text_tokenizer(num_words = max_words)
# tokenizer %>% keras::fit_text_tokenizer(movie_review$review)
```

**14. Transform each text into a sequence of integers (word indices) and use the `pad_sequences` function to pad the sequences.**

```{r}
# transform each text
sequences <- tokenizer %>% keras::texts_to_sequences(movie_review$review)

# pad the sequences
sequences <- pad_sequences(sequences, maxlen = maxlen)
```

**15. Convert the sequence into a dataframe.**

```{r}
# convert the sequence into a dataframe
dic <- as.data.frame(sequences)
```

**16. Use the code below to join the dataframe of sequences (word indices) from the IMDB reviews with GloVe pre-trained word vectors.**

```{r}
# join the words with GloVe vectors and
# if a word does not exist in GloVe, then fill NA's with 0
word_embeds <- dic %>%
  left_join(vectors) %>%
  select(starts_with("dim")) %>%
  replace(., is.na(.), 0) %>%
  as.matrix()
## Joining with `by = join_by(word)`
```

**17. Extract the outcome variable from the `sentiment` column in the original dataframe and name it `y_train.`**

**18. Use the `Keras` functional API and create a recurrent neural network model as below. Can you describe this model?**

```{r}
# Use Keras Functional API
input <- layer_input(shape = list(maxlen), name = "input")

model <- input %>%
  layer_embedding(input_dim = max_words, output_dim = dim_size, input_length = maxlen,
                  # put weights into list and do not allow training
                  weights = list(word_embeds), trainable = FALSE) %>%
  layer_spatial_dropout_1d(rate = 0.2) %>%
  bidirectional(
    layer_gru(units = 80, return_sequences = TRUE)
  )
max_pool <- model %>% layer_global_max_pooling_1d()
ave_pool <- model %>% layer_global_average_pooling_1d()

output <- layer_concatenate(list(ave_pool, max_pool)) %>%
  layer_dense(units = 1, activation = "sigmoid")

model <- keras_model(input, output)

# model summary
model
```

**19. Compile the model with an ‘adam’ optimizer, and the binary_crossentropy loss. You can choose accuracy or AUC for the metrics.**

**20. Fit the model with 10 epochs (iterations), batch_size = 32, and validation_split = 0.2. Check the training performance versus the validation performance.**
